A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-07-16 paper/paper.tex
paper: fix some obvious issues in last few sections

\section{Introduction}
% - semisimplicial/semicubical sets in indexed form
%   - clarify indexed/fibered, justify advantages of the respective choice
%     - the presentation matters when expressed in indexed way!
%   - compare to indexed VV, previous indexed HH, indexed Pang-Luo
% - iterated unary/binary parametricity has a semi-simplicial/semi-cubical structure
% - nu-types as a uniform definition of semi-simplicial/semi-cubical
%   iterated parametricity gives an alternative presentation of indexed semi-simplicial/semi-cubical
%   this is what we described in the paper
% - perspective: models of cubical type theory using indexed cubical sets, hence preserving more definitial rules
% - mention also the semi-simplicial types problem

\subsection*{Fibered vs indexed presentation of semi-simplicial and semi-cubical sets}
A family of sets can commonly be represented in two ways: as a family properly speaking, indexed over a given set $S$ or as a set $T$ together with a map specifying for each element of the set $T$ on which index in $S$ it depends. In the first case, we call it an \emph{indexed} presentation. In the latter case, the set associated to a given index is the fiber, i.e. inverse image, of the index under the map, so we call it a \emph{fibered} presentation. The two presentations are equivalent and the equivalence can be phrased concisely in the language of homotopy type theory~\cite{hottbook} as a fibered/indexed correspondence\footnote{At a discussion on the Zulip channel for the HoTT community, alternative names fibration/family correspondence and unbundled/bundled correspondence were proposed, among others. In our experience, the opposition fibered/indexed is spontaneously understood by many listeners. In particular, it echoes to the Grothendieck construction of fibered categories from indexed categories. The most elementary instance of the correspondence, with $\Type$ instead of $\U$, that is $(\Sigma S: \Type. (S \rightarrow T)) \simeq (T \rightarrow \Type)$, is sometimes called ``Grothendieck construction for dummies'' (its proof requires univalence, see~\cite{hottbook}).}
\begin{equation*}
  \mbox{(fibered)}\qquad(\Sigma S: \U. (S \rightarrow T)) ~\simeq~ (T \rightarrow \U)\qquad \mbox{(indexed)}
\end{equation*}
where $\U$ represents sets in homotopy type theory as the subset of types with a unique proof of equality (within some universe).

@@

\begin{align*}
  X_0 & :              & \U                                                     \\
  X_1 & :              & X_0 \rightarrow X_0 \rightarrow \U                     \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \rightarrow X_1(a)(c) \rightarrow X_1 (b)(d) \\
      &                & \rightarrow X_1 (c)(d) \rightarrow \U                  \\
  \ldots
\end{align*}

The idea for such indexed presentation of presheaves over a direct category was mentioned at the Univalent Foundations year in the context of defining semi-simplicial types~\footnote{See e.g. \url{https://ncatlab.org/nlab/show/semi-simplicial+types+in+homotopy+type+theory}.}. A couple of constructions were proposed since then. One, by \cite{voevodsky12}, relies on the presentation of semi-simplicial sets as a presheaf over increasing injective maps between finite ordinals. Another, by \cite{herbelin15}\footnote{With hindsight, the title of the paper ``A dependently-typed construction of semi-simplicial types'' is somehow confusing: it was part of a project at defining semi-simplicial types, but, in practice, the definition is done in a type theory with Unicity of Identity Proofs (UIP), meaning in a type theory where types (intended to represent arbitrary $\infty$-groupoids) are actually sets ($0$-truncated $\infty$-groupoids, that is, morally, $0$-groupoids). The confusion was however common at that time.}, and fully formalized in Coq, relies on the presentation of semi-simplicial sets as a presheaf over face maps. Another by~\cite{part15}, formalized in (an emulation of logic-enriched homotopy type theory in) the Plastic proof assistant, and yet another by \cite{annenkovCK17}, formalized in (an emulation of a two-level type theory in) the Agda proof assistant~\footnote{\url{https://github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}}, also rely on the presentation of the semi-simplicial category from increasing injective functions over finite ordinals.

@@

In category-theoretic terms, the indexing set of the family $X_n$ in the indexed presentation is called a \emph{matching object}. The indexed presentation of presheaves over a direct category can then be identified with a subclass of ordinary presheaves known as Reedy fibrant where Reedy fibrancy precisely expresses that the set $X_n$ in the fibered presentation is a $\Sigma$-type over an instance up to isomorphism of the matching object at level $n$~\citep{shulman14,kraus17}.

@@

\section{Semi-simplicial and semi-cubical sets\label{sec:nu}}
In this section, we generalize semi-simplicial and semi-cubical sets to, what we call, $\nu$-sets, subsuming the earlier definitions. We start by first explicitly working out the cases of semi-simplicial and semi-cubical sets. %, in a way that is similar to a previous art~(\citeyear{aitchison86}).

@@

\section{Relating to parametricity\label{sec:rel-param}}
The process of construction of the type of $X_1$ from that of $X_0$, and from the type of $X_2$ to that of $X_1$ in the last section, is similar to applying a binary parametricity translation and expecting the resulting translation to be inhabited. The parametricity which we consider interprets a type $A$ by a graph $A_\kstar$ over this type, and a term $t: A$ as an edge in $A_\kstar(t)(t)$. In particular, $\U$ is interpreted as the graph ${\U}_\kstar$, which takes $A_L$ and $A_R$ in $\U$ and returns the type $A_L \rightarrow A_R \rightarrow \U$ of graphs over $A_L$ and $A_R$. Also, for $A$ interpreted by $A_\kstar$ and $B$ interpreted by $B_\kstar$, a dependent function type $\Pi a: A.\, B$ is interpreted as the graph $(\Pi a: A.\, B)_\kstar$ that takes two functions $f_L$ and $f_R$ of type $\Pi a: A.\, B$, and expresses that these functions map related arguments in $A$ to related arguments in $B$:
\begin{align*}
   & (\Pi a: A.\, B)_\kstar(f_L)(f_R) \defeq                                                      \\
   & \Pi a_L: A.\, \Pi a_R: A.\, \Pi a_\kstar: A_\kstar(a_L)(a_R).\, B_\kstar(f_L(a_L))(f_R(a_R))
\end{align*}

@@

In Section~\ref{sec:ett} and Tables~\ref{tab:coind}, \ref{tab:core}, \ref{tab:frames}, \ref{tab:faces}, \ref{tab:coh}, we give an informal, mathematical definition in Extensional Type Theory, which roughly corresponds to a level of reasoning ``à la set theory'' where equalities are thought as ``holding on the nose'' without having to mention them explicitly once they are justified. In Section~\ref{sec:itt}, we give the same mathematical definition in Intensional Type Theory, where transporting along coherence conditions is now made explicit. Adding this level of details helps to understand the specificities of coherence conditions.

But first, we start with a section gradually and informally explaining the intuition of the construction. It can be read in parallel with the tables if wished.

@@

For the presentation corresponding to the formalization in \emph{intensional} type theory, see \ref{sec:itt}.

@@

Note that for a fixed constant $n$, the coherence conditions evaluate to a reflexivity proof, so that the construction evaluates to an effective sequence of types of iterated relations not mentioning them anymore. For instance, it produces Fig.~\ref{fig:intuition} for $n=2$ and $\nu=2$.

@@

We now present the definition in \emph{intensional} type theory~\cite{martinlof84,coq22}, making explicit the need for transport along coherence proofs.

@@

The way recursion is implemented is still left implicit at this stage. See \ref{sec:mech} for the details.

Depending on how inequality on natural numbers is defined, different equalities may hold or not definitionally, such as transitivity of inequality or the computational properties of induction over inequality proofs. Details are given in~\ref{sec:mech}.

@@

Here $\SProp$ is a definitionally proof-irrelevant impredicative universe at the bottom of the universe hierarchy~\cite{gilbert19}. By placing the definition in \SProp, we have definitional equality of all inequality proofs. For the purpose of unification, however, this definition does not go far enough. Consider the unification problems:

@@

\section{Future work}
In the cubical case, we expect the construction to eventually provide a model of (some version of) parametric type theory~\cite{nuyts17,cavallo19} by adding degeneracies, a hierarchy of universes (as sketched e.g. in a talk by Herbelin at the HoTT-UF workshop for the bridge case~\cite{herbelin-hott-uf}), as well as reasoning modulo permutations~\cite{grandis03}.

By equipping the universe construction with a structure of equivalences, as suggested along the lines of Altenkirch and Kaposi~\cite{altenkirch15}, we expect the construction to be able to serve as a basis for syntactic models of various versions of cubical type theory~\cite{bezem13,cohen16,angiuli21}, saving the detour via the fibered approach inherent to usual presheaf models. This would a priori preserve definitional properties which may be lost when detouring via presheaves. In particular, we conjecture being able to justify univalence holding definitionally. Our approach would also definitively ground cubical type theory in iterated parametricity.

>>>>

\section{Introduction}
\subsection*{Fibered vs indexed presentation of semi-simplicial and semi-cubical sets}
A family of sets can commonly be represented in two ways: as a family properly speaking, indexed over a given set $S$ or as a set $T$ together with a map specifying for each element of the set $T$ on which index in $S$ it depends. In the first case, we call it an \emph{indexed} presentation. In the latter case, the set associated to a given index is the fiber, i.e. inverse image, of the index under the map, so we call it a \emph{fibered} presentation. The two presentations are equivalent and the equivalence can be phrased concisely in the language of homotopy type theory~\cite{hottbook} as a fibered/indexed correspondence\footnote{At a discussion on the Zulip channel for the HoTT community, alternative names fibration/family correspondence and unbundled/bundled correspondence were proposed, among others. In our experience, the opposition fibered/indexed is spontaneously understood by many listeners. In particular, it echoes to the Grothendieck construction of fibered categories from indexed categories. The most elementary instance of the correspondence, with $\Type$ instead of $\U$, that is $(\Sigma S: \Type. (S \rightarrow T)) \simeq (T \rightarrow \Type)$, is sometimes called ``Grothendieck construction for dummies'' (its proof requires univalence, see~\cite{hottbook}).}
\begin{equation*}
  \mbox{(fibered)}\qquad(\Sigma S: \U. (S \rightarrow T)) ~\simeq~ (T \rightarrow \U)\qquad \mbox{(indexed)}
\end{equation*}
where $\U$ represents in homotopy type theory the subset of types with a unique proof of equality (within some universe).

@@

\begin{align*}
  X_0 & :              & \U                                                                                           \\
  X_1 & :              & X_0 \rightarrow X_0 \rightarrow \U                                                           \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \rightarrow X_1(a)(c) \rightarrow X_1 (b)(d) \rightarrow X_1 (c)(d) \rightarrow \U \\
  \ldots
\end{align*}

The idea for such indexed presentation of presheaves over a direct category was mentioned at the Univalent Foundations year in the context of defining semi-simplicial types~\footnote{See e.g. \url{https://ncatlab.org/nlab/show/semi-simplicial+types+in+homotopy+type+theory}.}. A couple of constructions were proposed since then. One, by \cite{voevodsky12}, relies on the presentation of semi-simplicial sets as a presheaf over increasing injective maps between finite ordinals. Another, by \cite{herbelin15}\footnote{In hindsight, the title of the paper ``A dependently-typed construction of semi-simplicial types'' is somewhat confusing: it was part of a project for defining semi-simplicial types, but, in practice, the definition is done in a type theory with Unicity of Identity Proofs (UIP), meaning in a type theory where types (intended to represent arbitrary $\infty$-groupoids) are actually sets ($0$-truncated $\infty$-groupoids, that is, morally, $0$-groupoids). The confusion was however common at the time.}, and fully formalized in Coq, relies on the presentation of semi-simplicial sets as a presheaf over face maps. Another by~\cite{part15}, formalized in (an emulation of logic-enriched homotopy type theory in) the Plastic proof assistant, and yet another by \cite{annenkovCK17}, formalized in (an emulation of a two-level type theory in) the Agda proof assistant~\footnote{\url{https://github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}}, also rely on the presentation of the semi-simplicial category from increasing injective functions over finite ordinals.

@@

In category-theoretic terms, the indexing set of the family $X_n$ in the indexed presentation is the \emph{matching object} of the presheaf at $n$. The indexed presentation of presheaves over a direct category can then be identified with a subclass of ordinary presheaves known as Reedy fibrant where Reedy fibrancy precisely expresses that the set $X_n$ in the fibered presentation is a $\Sigma$-type over an instance up to isomorphism of the matching object at level $n$~\citep{shulman14,kraus17}.

@@

\section{Semi-simplicial and semi-cubical sets\label{sec:nu}}
In this section, we generalize semi-simplicial and semi-cubical sets to, what we call, $\nu$-sets, subsuming the earlier definitions. We start by first explicitly working out the cases of semi-simplicial and semi-cubical sets.

@@

\section{Relating to parametricity\label{sec:rel-param}}
Let us recall from the introduction, the form taken by the indexed presentation of a semi-cubical set:
\begin{align*}
  X_0 & :              & \U                                                                                           \\
  X_1 & :              & X_0 \rightarrow X_0 \rightarrow \U                                                           \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \rightarrow X_1(a)(c) \rightarrow X_1 (b)(d) \rightarrow X_1 (c)(d) \rightarrow \U \\
  \ldots
\end{align*}

In there, the process of construction of the type of $X_1$ from that of $X_0$, and from the type of $X_2$ to that of $X_1$, and so on, is similar to iteratively applying a binary parametricity translation. The parametricity which we consider interprets a type $A$ by a graph $A_\kstar$ over this type, and a term $t: A$ as an edge in $A_\kstar(t)(t)$. In particular, $\U$ is interpreted as the graph ${\U}_\kstar$, which takes $A_L$ and $A_R$ in $\U$ and returns the type $A_L \rightarrow A_R \rightarrow \U$ of graphs over $A_L$ and $A_R$. Also, for $A$ interpreted by $A_\kstar$ and $B$ interpreted by $B_\kstar$, a dependent function type $\Pi a: A.\, B$ is interpreted as the graph $(\Pi a: A.\, B)_\kstar$ that takes two functions $f_L$ and $f_R$ of type $\Pi a: A.\, B$, and expresses that these functions map related arguments in $A$ to related arguments in $B$:
\begin{align*}
   & (\Pi a: A.\, B)_\kstar(f_L)(f_R) \defeq                                                      \\
   & \Pi a_L: A.\, \Pi a_R: A.\, \Pi a_\kstar: A_\kstar(a_L)(a_R).\, B_\kstar(f_L(a_L))(f_R(a_R))
\end{align*}

@@

In tables \ref{tab:coind}, \ref{tab:core}, \ref{tab:frames}, \ref{tab:faces}, \ref{tab:coh}, we give an informal, mathematical definition in Extensional Type Theory, which roughly corresponds to a level of reasoning ``à la set theory'' where equalities are thought as ``holding on the nose'' without having to mention them explicitly once they are justified. In section \ref{sec:itt}, we give the same mathematical definition in Intensional Type Theory, where transporting along coherence conditions is now made explicit. Adding this level of details helps to understand the specificities of coherence conditions.

@@

@@

Note that for a fixed constant $n$, the coherence conditions evaluate to a reflexivity proof, so that the construction evaluates to an effective sequence of types of iterated relations not mentioning them anymore.

@@

We now present the definition in \emph{intensional} type theory~\citep{martinlof84,coq22}, making explicit the need for transport along coherence proofs.

@@

The way recursion is implemented is still left implicit at this stage. Also, depending on how inequality on natural numbers is defined, different equalities may hold or not definitionally, such as transitivity of inequality or the computational properties of induction over inequality proofs. Details are given in \ref{sec:mech}.

@@

Here $\SProp$ is a definitionally proof-irrelevant impredicative universe at the bottom of the universe hierarchy~\citep{gilbert19}. By placing the definition in \SProp, we have definitional equality of all inequality proofs. For the purpose of unification, however, this definition does not go far enough. Consider the unification problems:

@@

\section{Future work}
In the cubical case, we expect the construction to eventually provide a model of (some version of) parametric type theory~\citep{nuyts17,cavallo19} by adding degeneracies, a hierarchy of universes (as sketched e.g. in a talk by Herbelin at the HoTT-UF workshop for the bridge case~(\citeyear{herbelin-hott-uf})), as well as reasoning modulo permutations~\citep{grandis03}.

By equipping the universe construction with a structure of equivalences, as suggested along the lines of Altenkirch and Kaposi~(\citeyear{altenkirch15}), we expect the construction to be able to serve as a basis for syntactic models of various versions of cubical type theory~\citep{bezem13,cohen16,angiuli21}, saving the detour via the fibered approach inherent to usual presheaf models. This would a priori preserve definitional properties which may be lost when detouring via presheaves. In particular, we conjecture being able to justify univalence holding definitionally. Our approach would also definitively ground cubical type theory in iterated parametricity.
